---
title: "Analyzing Sociodemographic Inequities in Smartphone Usage and Privacy Behaviors"
author: "Chriss Jordan Oboa"
date: "2024-12-02"
output:
  pdf_document: default
  word_document: default
latex_engine: xelatex
geometry: margin=1in
fontsize: 12pt
---

#Libraries

```{r setup-libraries}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(broom)
library(tinytex)
```

#Step 0: Data Inspection
```{r setup-dataset}
# Load the dataset (adjust the path to where your dataset is stored)
Data_Original <- read.csv("~/Downloads/DataAndSociety_DataAccessGrant_SurveyDataFiles-2/Survey Data Files/Data_Original.csv")


#duplicate the data
Data <- Data_Original

# Inspecting the structure of the dataset
str(Data)

# Viewing the first few rows
head(Data)

# Checking for summary statistics and missing values
summary(Data)
```

#Step1

##Step 1a: Gender Inequities in Smartphone Usage

###Data Preparation
```{r }
# Recode variables
Data <- Data %>%
  mutate(
    sex_clean = ifelse(sex == 1, "Male", "Female"),  # Recode `sex`
    smartphone_primary = ifelse(SmartPhone == 1, TRUE, FALSE)  # Recode `SmartPhone`
  )

# Check recoded variables
table(Data$sex_clean, useNA = "ifany")
table(Data$smartphone_primary, useNA = "ifany")
```

###Chi-Square Test for Gender Inequities
```{r }
# Filtering for valid responses
Data_clean <- Data %>%
  filter(!is.na(sex_clean) & !is.na(smartphone_primary))

# Creating a contingency table
gender_smartphone_table <- table(Data_clean$sex_clean, Data_clean$smartphone_primary)

# Performing chi-square test
chi_test_gender <- chisq.test(gender_smartphone_table)

# Display results
print(gender_smartphone_table)
print(chi_test_gender)
```

Males using smartphones primarily
```{r }
male_percentage <- gender_smartphone_table["Male", "TRUE"] / sum(gender_smartphone_table["Male", ]) * 100
print(male_percentage)
```

Females using smartphones primarily
```{r }
female_percentage <- gender_smartphone_table["Female", "TRUE"] / sum(gender_smartphone_table["Female", ]) * 100
print(female_percentage)
```

###Interpretation
The chi-square test for gender indicated a statistically significant association between gender and smartphone usage. Among respondents, it shows that 62.5% of males used smartphones primarily compared to 55.9% of females.

##Step 1b: Income Inequities in Smartphone Usage

###Data Preparation
```{r }
# Recoding the income variable
Data <- Data %>%
  mutate(
    income_clean = ifelse(inc %in% c(99, 98, 9), NA, inc),  # Recode non-responses
    low_income = case_when(
      income_clean %in% c(1, 2, 3, 4) ~ TRUE,   # Categories 1–4 represent low income
      income_clean %in% c(5, 6, 7, 8) ~ FALSE, # Categories 5–8 represent higher income
      TRUE ~ NA                              # Handle missing or non-responses
    )
  )

# Check the recoded variable
table(Data$low_income, useNA = "ifany")

# Verify the recoded `income_clean` variable
summary(Data$income_clean)
table(Data$income_clean, useNA = "ifany")
```

###Chi-Square Test for Income Inequities
```{r }
# Filtering valid responses
Data_clean <- Data %>%
  filter(!is.na(low_income) & !is.na(smartphone_primary))

# Create a contingency table
income_smartphone_table <- table(Data_clean$low_income, Data_clean$smartphone_primary)

# Perform chi-square test
chi_test_income <- chisq.test(income_smartphone_table)

# Displaying results
print(income_smartphone_table)
print(chi_test_income)
```

Low-income respondents using smartphones primarily
```{r }
low_income_percentage <- income_smartphone_table["TRUE", "TRUE"] / sum(income_smartphone_table["TRUE", ]) * 100
print(low_income_percentage)
```

Higher-income respondents using smartphones primarily
```{r }
high_income_percentage <- income_smartphone_table["FALSE", "TRUE"] / sum(income_smartphone_table["FALSE", ]) * 100
print(high_income_percentage)
```

###Interpretation
The chi-square test for income showed a significant association between income and smartphone usage. Among low-income respondents, 45.7% reported using smartphones as their primary internet device, compared to 75.2% of higher-income respondents.


#Step 2: Sociodemographic Inequities in Privacy/Security Behaviors

##Logistic Regression Model

###Data Preparation
```{r }
# Recode sociodemographic variables
Data <- Data %>%
  mutate(
    raceethn_clean = case_when(
      racethn == 1 ~ "White",
      racethn == 2 ~ "Black",
      racethn == 3 ~ "Hispanic",
      racethn == 4 ~ "Other",
      racethn %in% c(9, 99) ~ NA_character_
    ),
    education_clean = case_when(
      educ2 %in% c(1, 2) ~ "No College",
      educ2 %in% c(3, 4, 5) ~ "Some College",
      educ2 %in% c(6, 7) ~ "College Graduate",
      TRUE ~ NA_character_
    ),
    sex_clean = ifelse(sex == 1, "Male", "Female"),
    low_income = ifelse(income_clean < 40000, TRUE, FALSE)  # Already recoded
  )

# Creating a binary variable for SM1 behaviors
Data <- Data %>%
  mutate(
    privacy_behavior = ifelse(rowSums(select(., starts_with("sm1")), na.rm = TRUE) > 0, TRUE, FALSE)
  )

# Filter complete cases
Data_clean <- Data %>%
  filter(
    !is.na(raceethn_clean),
    !is.na(low_income),
    !is.na(education_clean),
    !is.na(sex_clean),
    !is.na(privacy_behavior)
  )
```

###Logistic Regression Model
```{r }
# Fitting logistic regression model
privacy_model <- glm(
  privacy_behavior ~ low_income + sex_clean + raceethn_clean + education_clean + age,
  data = Data_clean,
  family = binomial
)

# View regression summary
summary(privacy_model)
```

###Odds Ratios and Visualization
```{r }
# Excluding the NA coefficients 
odds_ratios <- exp(coef(privacy_model)[!is.na(coef(privacy_model))])
conf_intervals <- exp(confint(privacy_model)[!is.na(coef(privacy_model)), ])

# Combining results into a table
results <- data.frame(
  Variable = names(odds_ratios),
  Odds_Ratio = odds_ratios,
  Lower_CI = conf_intervals[, 1],
  Upper_CI = conf_intervals[, 2],
  p_value = coef(summary(privacy_model))[, 4][!is.na(coef(privacy_model))]
)

# Display the results
print(results)
```


### Bar Plot for Odds Ratios
```{r }
#Bar Plot for Odds Ratios
categorical_effects <- results %>%
  filter(!Variable %in% c("(Intercept)", "age"))  # Exclude intercept and age for the bar plot

ggplot(categorical_effects, aes(x = reorder(Variable, Odds_Ratio), y = Odds_Ratio)) +
  geom_bar(stat = "identity", fill = "skyblue", color = "black") +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width = 0.3, color = "red") +
  scale_y_log10() +  # Log scale for odds ratios
  labs(
    title = "Odds Ratios for Sociodemographic Variables",
    subtitle = "Log scale used to highlight differences in effect size",
    x = "Sociodemographic Variables",
    y = "Odds Ratio (Log Scale)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, color = "darkblue", size = 12),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 14, face = "italic")
  )
  
```


### Line Plot for Age Effect
```{r }
#Line Plot for Age Effect
age_range <- seq(18, 80, 1)
predicted_odds <- exp(log(odds_ratios["age"]) * age_range)  # Simulate odds based on age

# Data frame for plotting
age_effect <- data.frame(Age = age_range, Odds = predicted_odds)

ggplot(age_effect, aes(x = Age, y = Odds)) +
  geom_line(color = "darkred", size = 1.2) +
  geom_point(data = age_effect[seq(1, nrow(age_effect), by = 5), ], aes(x = Age, y = Odds), size = 2, color = "red") +
  labs(
    title = "Effect of Age on Privacy Behaviors",
    subtitle = "Odds of engaging in privacy behaviors decrease with age",
    x = "Age (Years)",
    y = "Odds of Privacy Behaviors"
  ) +
  annotate("text", x = 60, y = max(age_effect$Odds), label = "Decreasing trend", color = "darkred", hjust = 1.2) +
  theme_minimal() +
  theme(
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 14, face = "italic")
  )
```

###Key Findings:

Race/Ethnicity: White respondents were significantly more likely to engage in privacy/security behaviors compared to Black and Hispanic respondents.

Education: Respondents with no college education were significantly less likely to engage in privacy/security behaviors.

Age: Older individuals showed a slight reduction in privacy behavior engagement with increasing age.

#Step 3: Justify Your Analysis Choices
##In 250 words or less, explain the reasoning behind your approach for each RQ.

RQ1a: Gender Inequities in Smartphone Usage
We used a chi-square test to explore the relationship between gender and smartphone usage because both variables are categorical. The variable sex was recoded into sex_clean with values "Male" and "Female." Similarly, the SmartPhone variable was recoded into smartphone_primary as a binary indicator of whether respondents primarily used their smartphones for internet access. Non-responses were excluded to ensure valid comparisons. The chi-square test was chosen to test independence between gender and smartphone usage.

RQ1b: Income Inequities in Smartphone Usage
The income variable was transformed into low_income, a binary indicator for respondents earning below $40,000 annually, to focus on disparities between low and higher-income groups. Non-responses were excluded. A chi-square test was again used to examine the association between income status and smartphone usage.

RQ2: Sociodemographic Inequities in Privacy/Security Behaviors
Logistic regression was employed to investigate sociodemographic predictors of privacy/security behaviors (privacy_behavior). Predictors included income, gender, race/ethnicity, education, and age. Categorical variables (race, education) were collapsed into fewer levels to ensure sufficient sample sizes for comparisons. Non-responses were also excluded. Logistic regression was selected because it accommodates a binary dependent variable (privacy behavior: yes/no).


###Effect of Education and Age
```{r }
# Filtering results for categorical variables
categorical_effects <- results %>%
  filter(!Variable %in% c("(Intercept)", "age"))  # Exclude intercept and age for the bar plot
```

```{r }
ggplot(categorical_effects, aes(x = Variable, y = Odds_Ratio)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width = 0.2) +
  scale_y_log10() +  # Log scale for odds ratios
  theme_minimal() +
  labs(
    title = "Odds Ratios for Sociodemographic Factors Affecting Privacy Behaviors",
    subtitle = "Visualizing the impact of education, race, and gender",
    x = "Sociodemographic Variables",
    y = "Odds Ratio (Log Scale)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

###Effect of Age
```{r }
# Generating data for predicted age effect
age_range <- seq(18, 80, 1)  # Age range for prediction
predicted_odds <- exp(log(odds_ratios["age"]) * age_range)  # Simulate odds based on age

# Create a data frame for plotting
age_effect <- data.frame(Age = age_range, Odds = predicted_odds)
```

```{r }
# Create a data frame for plotting
age_effect <- data.frame(Age = age_range, Odds = predicted_odds)

# Plotting the effect of age
ggplot(age_effect, aes(x = Age, y = Odds)) +
  geom_line(color = "darkred", linewidth = 1) +  # Use linewidth instead of size
  theme_minimal() +
  labs(
    title = "Effect of Age on Privacy Behaviors",
    subtitle = "Simulated odds for increasing age",
    x = "Age (Years)",
    y = "Odds of Engaging in Privacy Behaviors"
  ) +
  annotate("text", x = 60, y = max(age_effect$Odds), label = "Decreasing trend", color = "darkred", hjust = 1)
```

#Step 4: Reporting Your Results (400 Words or Less)

##RQ1a: Gender Inequities in Smartphone Usage
To investigate gender differences in smartphone usage, we conducted a chi-square test using recoded gender (sex_clean) and smartphone usage (smartphone_primary) variables. Results showed a statistically significant association between gender and smartphone usage (x² = 13.1, p < 0.001). About 62.5% of males used smartphones primarily for internet access, compared to 55.9% of females. This disparity highlights that males are more reliant on smartphones for internet access than females.

##RQ1b: Income Inequities in Smartphone Usage
Income's impact on smartphone usage was analyzed using the binary variable low_income (< $40,000). The chi-square test revealed a strong association. While 45.7% of low-income respondents relied on smartphones primarily for internet access, this figure was significantly higher for higher-income respondents (75.2%). These results emphasize income as a critical factor in smartphone dependency.

##RQ2: Sociodemographic Inequities in Privacy/Security Behaviors
We used logistic regression to explore predictors of privacy/security behaviors (privacy_behavior). 

Key findings included:

Race/Ethnicity: White respondents were significantly more likely to engage in privacy/security behaviors than Black or Hispanic respondents (OR = 1.40, p = 0.027).

Education: Education was a strong predictor. Respondents without college education were 86% less likely to engage in privacy/security behaviors (OR = 0.14, p < 0.001). Those with some college education were also less likely (OR = 0.40, p < 0.001).

Age: Older individuals showed a slight decline in engagement, with a 6% reduction in odds per additional year (OR = 0.94, p < 0.001).

Gender and Income: Neither was significant after controlling for other factors.

These findings underscore the influence of race, education, and age on privacy behavior, while gender and income exhibited no independent effects.

#Extra Credit: Regional Differences in Privacy/Security Behaviors
We examined regional differences in privacy/security behaviors using logistic regression with region_clean as the predictor.

##Logistic Regression by Region
# Recode region variable
```{r }
# Recode region variable
Data <- Data %>%
  mutate(region_clean = case_when(
    scregion == 1 ~ "Northeast",
    scregion == 2 ~ "Midwest",
    scregion == 3 ~ "South",
    scregion == 4 ~ "West",
    TRUE ~ NA_character_
  ))

# Convert region_clean to a factor and set "Northeast" as the reference level
Data <- Data %>%
  mutate(region_clean = factor(region_clean, levels = c("Northeast", "Midwest", "South", "West")))

# Filter complete cases for logistic regression
Data_clean <- Data %>%
  filter(
    !is.na(raceethn_clean),
    !is.na(low_income),
    !is.na(education_clean),
    !is.na(sex_clean),
    !is.na(privacy_behavior),
    !is.na(region_clean)  # Ensure no missing values for region_clean
  )

# Fit logistic regression model with "Northeast" as the reference level
region_model <- glm(privacy_behavior ~ region_clean, data = Data_clean, family = binomial)

# View regression summary
summary(region_model)

```
```{r }
# Set "Northeast" as the reference level
Data_clean$region_clean <- relevel(Data_clean$region_clean, ref = "Northeast")


```
##Visualizing Regional Effects
```{r }
# Extract odds ratios and confidence intervals
odds_ratios_region <- exp(coef(region_model))
conf_intervals_region <- exp(confint(region_model))

# Create a data frame for visualization
region_results <- data.frame(
  Region = names(odds_ratios_region),
  Odds_Ratio = odds_ratios_region,
  Lower_CI = conf_intervals_region[, 1],
  Upper_CI = conf_intervals_region[, 2]
)

# Exclude intercept from the results
region_results <- region_results[-1, ]  # Remove "(Intercept)"

# Add Northeast as the reference point (Odds Ratio = 1)
region_results <- region_results %>%
  add_row(
    Region = "Northeast",
    Odds_Ratio = 1,
    Lower_CI = 1,
    Upper_CI = 1
  ) %>%
  arrange(factor(Region, levels = c("Northeast", "Midwest", "South", "West")))

# Plot the regional odds ratios
ggplot(region_results, aes(x = Region, y = Odds_Ratio)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width = 0.2) +
  scale_y_log10() +  # Log scale for odds ratios
  labs(
    title = "Regional Effects on Privacy Behaviors",
    subtitle = "Odds ratios compared to the reference region (Northeast)",
    x = "Region",
    y = "Odds Ratio (Log Scale)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, face = "italic")
  )
```

